---
title: "CITE-seq optimization - ADT in cells vs empty drops"
author: "Terkild Brink Buus"
date: "30/3/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
options(stringsAsFactors=FALSE)
```

## Load libraries etc.

```{r}
set.seed(114)
require("Seurat", quietly=T)
require("tidyverse", quietly=T)
require("Matrix", quietly=T)
require("DropletUtils", quietly=T)

## Load ggplot theme and defaults
source("R/ggplot_settings.R")

## Load helper functions (ggplot themes, biexp transformation etc.)
source("R/Utilities.R")

## Load predefined color schemes
source("R/color.R")

data.drive <- "F:"
outdir <- "C:/Users/Terkild/OneDrive - Københavns Universitet/Koralovlab/ECCITE-seq/20200106 Titration 1"
```

## Concentration bias in empty droplets?

Do antibodies used in high concentration dominate the empty drops?

ADT

```{r}
abpanel <- data.frame(readxl::read_excel("data/Supplementary_Table_1.xlsx"))
rownames(abpanel) <- abpanel$Marker


ADT.matrix <- kallisto.ADT
ADT.matrix.marker <- rownames(ADT.matrix)
ADT.matrix.conc <- abpanel[ADT.matrix.marker,"conc_µg_per_mL"]
markerUMI <- apply(ADT.matrix,1,sum)
markerUMI.inCell <- apply(ADT.matrix[,gex.aboveInf],1,sum)
markerUMI.inCell.freq <- markerUMI.inCell/sum(markerUMI.inCell)
markerUMI.inDrop <- markerUMI-markerUMI.inCell
markerUMI.inDrop.freq <- markerUMI.inDrop/sum(markerUMI.inDrop)

plotData <- rbind(data.frame(count=markerUMI.inCell, freq=markerUMI.inCell.freq, subset="Cell", marker=names(markerUMI.inCell.freq)), data.frame(count=markerUMI.inDrop, freq=markerUMI.inDrop.freq, subset="EmptyDrop", marker=names(markerUMI.inDrop.freq)))
plotData$conc <- abpanel[plotData$marker,"conc_µg_per_mL"]
plotData$marker <- factor(plotData$marker, levels=plotData$marker[order(plotData$conc[plotData$subset=="EmptyDrop"],plotData$freq[plotData$subset=="EmptyDrop"])])

scaleFUN <- function(x) sprintf("%.2f", x)
## MAYBE COMPARE ROW SUMS FROM READS vs UMI FROM CSC.

plotData <- plotData[order(plotData$marker, decreasing=TRUE),]
plotData$highlight <- 0
#plotData$highlight[which(plotData$marker == plotData$marker[plotData$freq >= 0.05] & plotData$subset == "EmptyDrop")]
highfreq <- which(plotData$marker %in% plotData$marker[plotData$freq >= 0.05])
plotData$highlight[c(highfreq)] <- 1
plotData$label <- plotData$marker
max.label <- plotData[plotData$freq >= 0.05,] %>% group_by(marker) %>% summarize(subset.max=subset[which.max(freq)])

plotData$label[(paste(plotData$marker,plotData$subset) %in% paste(max.label$marker,max.label$subset.max))==FALSE | plotData$freq < 0.05] <- NA

plotData$cummulativeFreq <- 0
plotData$cummulativeFreq[plotData$subset=="EmptyDrop"] <- cumsum(plotData$freq[plotData$subset=="EmptyDrop"])
plotData$cummulativeFreq[plotData$subset=="Cell"] <- cumsum(plotData$freq[plotData$subset=="Cell"])
p.gradient.hb.width <- 0.25

p.gradient.hb <- ggplot(plotData, aes(x=factor(as.character(subset), levels=c("EmptyDrop","Cell")), y=freq, fill=conc, group=marker)) + 
  geom_bar(stat="identity", position="stack", col=alpha("black",0.5), width=p.gradient.hb.width) + 
  scale_fill_viridis_c(trans="log2", labels=scaleFUN, breaks=c(0.0375,0.15,0.625,2.5,10)) + 
  scale_y_continuous(expand=c(0,0,0,0), position="right") + 
  scale_x_discrete(expand=c(0,0,0,0)) + 
  labs(fill="µg/mL") + 
  theme(panel.grid=element_blank(), axis.title.y=element_blank(), axis.text.x=element_text(angle=0, hjust=0.5, vjust=0.5)) + coord_flip(clip="off") + labs(y="UMI frequency (cells versus empty drops)")

p.gradient.hb2 <- p.gradient.hb + 
  geom_text(aes(y=cummulativeFreq-(freq/2), label=label), na.rm=TRUE, vjust=0.5, hjust=0.5,  angle=0, size=1.5, fontface="bold") + 
  theme(plot.margin=unit(c(0,0.2,0.2,0),"cm")) + 
  guides(alpha=FALSE)

plotData$xval <- (2-p.gradient.hb.width/2)
plotData$xval[plotData$subset == "EmptyDrop"] <- (1+p.gradient.hb.width/2)
plotData.poly2 <- plotData
plotData.poly2$cummulativeFreq <- plotData.poly2$cummulativeFreq-plotData.poly2$freq
plotData.poly <- rbind(plotData.poly2[order(plotData.poly2$marker, -plotData.poly2$xval, -plotData.poly2$cummulativeFreq, decreasing=TRUE),],plotData[order(plotData$marker, -plotData$xval, -plotData$cummulativeFreq, decreasing=FALSE),])

plotData.poly[plotData.poly$marker == "TCRgd",]

p.gradient.hb4 <- p.gradient.hb2 + 
  geom_polygon(data=plotData.poly, aes(x=xval,y=cummulativeFreq, group=marker, fill=conc, size=factor(highlight), col=factor(highlight)), alpha=0.5) + 
  scale_size_manual(values=c("0"=0.15, "1"=0.25)) + 
  scale_color_manual(values=c("0"=alpha("black",0.1), "1"=alpha("black",1)))

p.gradient.hb4

plotData$marker.rev <- factor(as.character(plotData$marker), levels=rev(levels(plotData$marker)))

p.barplot <- ggplot(plotData, aes(x=marker.rev, y=count/10^6, fill=subset)) + 
  geom_bar(stat="identity", position="dodge", color="black", width=0.65) + 
  geom_hline(yintercept=0, col="black") + 
  scale_fill_manual(values=c("black","lightgrey")) + 
  ggnewscale::new_scale_fill() + 
  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-0.050000,ymax=-0.010000,fill=conc), col="black") + 
  scale_fill_viridis_c(trans="log2", labels=scaleFUN, breaks=c(0.0375,0.15,0.625,2.5,10)) + 
  scale_x_discrete(expand=c(0, 0.5)) + 
  scale_y_continuous(expand=c(0.005,0.005,0,0)) + 
  facet_grid(~-conc, scales="free_x", space="free_x", switch="x") +
  labs(y=bquote("UMI count ("~10^6~")"), fill="µg/mL") + 
  theme(panel.border=element_blank(), 
        panel.grid=element_blank(), 
        panel.spacing=unit(0.5,"mm"),
        axis.line=element_line(), 
        axis.title.x=element_blank(), 
        strip.placement="outside", 
        strip.text=element_blank(), 
        legend.position=c(1,1), 
        legend.justification=c(1,1),
        #legend.direction="horizontal",
        legend.text.align=0, 
        legend.key.width=unit(0.3,"cm"), 
        legend.key.height=unit(0.4,"cm"), 
        legend.text=element_text(size=unit(5,"pt")))

#ragg::agg_png
png(file=file.path(outdir,"Figure 5.png"), width=figure.width.full, height=3.2, units=figure.unit, res=figure.resolution, antialias=figure.antialias)
cowplot::plot_grid(p.barplot, p.gradient.hb4 + theme(legend.position="none"), ncol=1, rel_heights=c(3.2,1), align="v", axis="lr", labels=c("A", "B"), label_size=panel.label_size, vjust=panel.label_vjust, hjust=panel.label_hjust)
dev.off()

```

HTO

```{r}
abpanel <- data.frame(readxl::read_excel("data/Supplementary_Table_1.xlsx"))
rownames(abpanel) <- abpanel$Marker

HTO

```{r}
abpanel <- data.frame(readxl::read_excel("data/Supplementary_Table_1.xlsx"))
rownames(abpanel) <- abpanel$Marker

markerUMI <- apply(kallisto.HTO,1,sum)
markerUMI.inCell <- apply(kallisto.HTO[,gex.aboveInf],1,sum)
markerUMI.inCell.freq <- markerUMI.inCell/sum(markerUMI.inCell)
markerUMI.inDrop <- markerUMI-markerUMI.inCell
markerUMI.inDrop.freq <- markerUMI.inDrop/sum(markerUMI.inDrop)

plotData <- rbind(data.frame(count=markerUMI.inCell, freq=markerUMI.inCell.freq, subset="Cell", marker=names(markerUMI.inCell.freq)), data.frame(count=markerUMI.inDrop, freq=markerUMI.inDrop.freq, subset="EmptyDrop", marker=names(markerUMI.inDrop.freq)))
plotData$conc <- 10
plotData$marker <- factor(plotData$marker)
plotData$marker <- factor(as.character(plotData$marker), levels=rev(levels(plotData$marker)))

scaleFUN <- function(x) sprintf("%.2f", x)

plotData <- plotData[order(plotData$marker, decreasing=TRUE),]
plotData$highlight <- 0
#plotData$highlight[which(plotData$marker == plotData$marker[plotData$freq >= 0.05] & plotData$subset == "EmptyDrop")]
highfreq <- which(plotData$marker %in% plotData$marker[plotData$freq >= 0.05])
plotData$highlight[c(highfreq)] <- 1
plotData$label <- plotData$marker
max.label <- plotData[plotData$freq >= 0.05,] %>% group_by(marker) %>% summarize(subset.max=subset[which.max(freq)])

plotData$label[(paste(plotData$marker,plotData$subset) %in% paste(max.label$marker,max.label$subset.max))==FALSE | plotData$freq < 0.05] <- NA

plotData$cummulativeFreq <- 0
plotData$cummulativeFreq[plotData$subset=="EmptyDrop"] <- cumsum(plotData$freq[plotData$subset=="EmptyDrop"])
plotData$cummulativeFreq[plotData$subset=="Cell"] <- cumsum(plotData$freq[plotData$subset=="Cell"])
p.gradient.hb.width <- 0.25

p.gradient.hb <- ggplot(plotData, aes(x=factor(as.character(subset), levels=c("EmptyDrop","Cell")), y=freq, fill=marker, group=marker)) + geom_bar(stat="identity", position="stack", col=alpha("black",0.5), width=p.gradient.hb.width) + scale_y_continuous(expand=c(0,0,0,0), position="right") + scale_x_discrete(expand=c(0,0,0,0)) + labs(fill="µg/mL") + theme(panel.grid=element_blank(), axis.title.y=element_blank(), axis.text.x=element_text(angle=0, hjust=0.5, vjust=0.5)) + coord_flip(clip="off") + labs(y="UMI frequency (cells versus empty drops)")
p.gradient.hb2 <- p.gradient.hb + geom_text(aes(y=cummulativeFreq-(freq/2), label=label), na.rm=TRUE, vjust=0.5, hjust=0.5,  angle=0, size=2.5, fontface="bold") + theme(plot.margin=unit(c(0,0.5,2,0),"cm")) + guides(alpha=FALSE)

plotData$xval <- (2-p.gradient.hb.width/2)
plotData$xval[plotData$subset == "EmptyDrop"] <- (1+p.gradient.hb.width/2)
plotData.poly2 <- plotData
plotData.poly2$cummulativeFreq <- plotData.poly2$cummulativeFreq-plotData.poly2$freq
plotData.poly <- rbind(plotData.poly2[order(plotData.poly2$marker, -plotData.poly2$xval, -plotData.poly2$cummulativeFreq, decreasing=TRUE),],plotData[order(plotData$marker, -plotData$xval, -plotData$cummulativeFreq, decreasing=FALSE),])

plotData.poly[plotData.poly$marker == "TCRgd",]

p.HTO.gradient.hb4 <- p.gradient.hb2 + geom_polygon(data=plotData.poly, aes(x=xval,y=cummulativeFreq, group=marker, fill=marker, size=factor(highlight), col=factor(highlight)), alpha=0.5) + scale_size_manual(values=c("0"=0.25, "1"=0.5)) + scale_color_manual(values=c("0"=alpha("black",0.1), "1"=alpha("black",1)))

p.HTO.gradient.hb4

plotData$marker.rev <- factor(as.character(plotData$marker), levels=rev(levels(plotData$marker)))

p.HTO.barplot <- ggplot(plotData, aes(x=marker.rev, y=count/10^6, fill=subset)) + geom_bar(stat="identity", position="dodge", color="black", width=0.65) + scale_x_discrete(expand=c(0, 0.5)) + scale_fill_manual(values=c("black","lightgrey")) + facet_grid(~marker.rev, scales="free_x", space="free_x", switch="x") + theme(panel.border=element_blank(), panel.grid=element_blank(), axis.line=element_line(), axis.title.x=element_blank(), panel.spacing=unit(0.1,"cm"), legend.position=c(1,1), legend.justification=c(1,1), strip.placement="outside", strip.text=element_blank(), legend.text.align=0) + scale_y_continuous(expand=c(0.005,0.005,0,0)) + ggnewscale::new_scale_fill() +  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-0.18000,ymax=-0.060000,fill=marker.rev), col="black") + geom_hline(yintercept=0, col="black") + labs(y=bquote("UMI count ("~10^6~")"), fill="µg/mL") + guides(fill=F)

#ragg::agg_png
png(filename=file.path(outdir,"EmptyDrops_HTO.png"), width=10, height=7, units = "in", res=300)
cowplot::plot_grid(p.barplot, p.HTO.barplot + theme(legend.position="none"), p.gradient.hb4 + theme(legend.position="none"), p.HTO.gradient.hb4 + theme(legend.position="none", axis.text.y=element_blank(), axis.title.x=element_blank()), ncol=2,  rel_heights=c(3,1), rel_widths=c(4,1), align="hv", axis="tblr", labels=c("A", "C", "B", "D"), hjust=0, vjust=1)
dev.off()

```

## COMPARE ROW SUMS FROM READS vs UMI FROM CSC.

Are reads evenly distributed among marker UMIs?

```{r}

dataCSCADTDir.reads <- "F:/Projects/ECCITE-seq/TotalSeqC_TitrationA/cite-seq-count/A1_S5_d1_ADT/read_count"

CSC.ADT.reads <- Read10X(data.dir=dataCSCADTDir.reads, gene.column=1)
CSC.ADT.reads <- CSC.ADT.reads[rownames(CSC.ADT.reads) != "unmapped",]

reads <- apply(CSC.ADT.reads,1,sum)
UMIs <- apply(CSC.ADT[,colnames(CSC.ADT.reads)],1,sum)

plotData <- data.frame(marker=gsub("(.*)\\-.*","\\1",names(reads)),reads=reads,UMIs=UMIs)

ggplot(plotData, aes(x=UMIs, y=reads)) + geom_point() + geom_smooth(method="lm") + scale_x_log10() + scale_y_log10()

reads.perCell <- as.data.frame(CSC.ADT.reads[,intersect(colnames(CSC.ADT.reads),gex.aboveInf)]) %>% mutate(marker=gsub("(.*)\\-.*","\\1",rownames(CSC.ADT.reads))) %>% pivot_longer(-marker)

UMIs.perCell <- as.data.frame(CSC.ADT[,intersect(colnames(CSC.ADT.reads),gex.aboveInf)]) %>% mutate(marker=gsub("(.*)\\-.*","\\1",rownames(CSC.ADT))) %>% pivot_longer(-marker)

plotData <- data.frame(marker=UMIs.perCell$marker,reads=reads.perCell$value,UMIs=UMIs.perCell$value)

ggplot(plotData, aes(x=UMIs, y=reads)) + geom_point(aes(color=abpanel[as.character(marker),"conc_µg_per_mL"])) + geom_smooth(method="lm") + scale_x_log10() + scale_y_log10() + scale_color_viridis_c(trans="log2", labels=scaleFUN, breaks=c(0.0375,0.15,0.625,2.5,10))

```




markerUMI <- apply(kallisto.HTO,1,sum)
markerUMI.inCell <- apply(kallisto.HTO[,gex.aboveInf],1,sum)
markerUMI.inCell.freq <- markerUMI.inCell/sum(markerUMI.inCell)
markerUMI.inDrop <- markerUMI-markerUMI.inCell
markerUMI.inDrop.freq <- markerUMI.inDrop/sum(markerUMI.inDrop)

plotData <- rbind(data.frame(count=markerUMI.inCell, freq=markerUMI.inCell.freq, subset="Cell", marker=names(markerUMI.inCell.freq)), data.frame(count=markerUMI.inDrop, freq=markerUMI.inDrop.freq, subset="EmptyDrop", marker=names(markerUMI.inDrop.freq)))
plotData$conc <- 10
plotData$marker <- factor(plotData$marker)
plotData$marker <- factor(as.character(plotData$marker), levels=rev(levels(plotData$marker)))

scaleFUN <- function(x) sprintf("%.2f", x)

plotData <- plotData[order(plotData$marker, decreasing=TRUE),]
plotData$highlight <- 0
#plotData$highlight[which(plotData$marker == plotData$marker[plotData$freq >= 0.05] & plotData$subset == "EmptyDrop")]
highfreq <- which(plotData$marker %in% plotData$marker[plotData$freq >= 0.05])
plotData$highlight[c(highfreq)] <- 1
plotData$label <- plotData$marker
max.label <- plotData[plotData$freq >= 0.05,] %>% group_by(marker) %>% summarize(subset.max=subset[which.max(freq)])

plotData$label[(paste(plotData$marker,plotData$subset) %in% paste(max.label$marker,max.label$subset.max))==FALSE | plotData$freq < 0.05] <- NA

plotData$cummulativeFreq <- 0
plotData$cummulativeFreq[plotData$subset=="EmptyDrop"] <- cumsum(plotData$freq[plotData$subset=="EmptyDrop"])
plotData$cummulativeFreq[plotData$subset=="Cell"] <- cumsum(plotData$freq[plotData$subset=="Cell"])
p.gradient.hb.width <- 0.25

p.gradient.hb <- ggplot(plotData, aes(x=factor(as.character(subset), levels=c("EmptyDrop","Cell")), y=freq, fill=marker, group=marker)) + geom_bar(stat="identity", position="stack", col=alpha("black",0.5), width=p.gradient.hb.width) + scale_y_continuous(expand=c(0,0,0,0), position="right") + scale_x_discrete(expand=c(0,0,0,0)) + labs(fill="µg/mL") + theme(panel.grid=element_blank(), axis.title.y=element_blank(), axis.text.x=element_text(angle=0, hjust=0.5, vjust=0.5)) + coord_flip(clip="off") + labs(y="UMI frequency (cells versus empty drops)")
p.gradient.hb2 <- p.gradient.hb + geom_text(aes(y=cummulativeFreq-(freq/2), label=label), na.rm=TRUE, vjust=0.5, hjust=0.5,  angle=0, size=2.5, fontface="bold") + theme(plot.margin=unit(c(0,0.5,2,0),"cm")) + guides(alpha=FALSE)

plotData$xval <- (2-p.gradient.hb.width/2)
plotData$xval[plotData$subset == "EmptyDrop"] <- (1+p.gradient.hb.width/2)
plotData.poly2 <- plotData
plotData.poly2$cummulativeFreq <- plotData.poly2$cummulativeFreq-plotData.poly2$freq
plotData.poly <- rbind(plotData.poly2[order(plotData.poly2$marker, -plotData.poly2$xval, -plotData.poly2$cummulativeFreq, decreasing=TRUE),],plotData[order(plotData$marker, -plotData$xval, -plotData$cummulativeFreq, decreasing=FALSE),])

plotData.poly[plotData.poly$marker == "TCRgd",]

p.HTO.gradient.hb4 <- p.gradient.hb2 + geom_polygon(data=plotData.poly, aes(x=xval,y=cummulativeFreq, group=marker, fill=marker, size=factor(highlight), col=factor(highlight)), alpha=0.5) + scale_size_manual(values=c("0"=0.25, "1"=0.5)) + scale_color_manual(values=c("0"=alpha("black",0.1), "1"=alpha("black",1)))

p.HTO.gradient.hb4

plotData$marker.rev <- factor(as.character(plotData$marker), levels=rev(levels(plotData$marker)))

p.HTO.barplot <- ggplot(plotData, aes(x=marker.rev, y=count/10^6, fill=subset)) + geom_bar(stat="identity", position="dodge", color="black", width=0.65) + scale_x_discrete(expand=c(0, 0.5)) + scale_fill_manual(values=c("black","lightgrey")) + facet_grid(~marker.rev, scales="free_x", space="free_x", switch="x") + theme(panel.border=element_blank(), panel.grid=element_blank(), axis.line=element_line(), axis.title.x=element_blank(), panel.spacing=unit(0.1,"cm"), legend.position=c(1,1), legend.justification=c(1,1), strip.placement="outside", strip.text=element_blank(), legend.text.align=0) + scale_y_continuous(expand=c(0.005,0.005,0,0)) + ggnewscale::new_scale_fill() +  geom_rect(aes(xmin=-Inf,xmax=Inf,ymin=-0.18000,ymax=-0.060000,fill=marker.rev), col="black") + geom_hline(yintercept=0, col="black") + labs(y=bquote("UMI count ("~10^6~")"), fill="µg/mL") + guides(fill=F)

#ragg::agg_png
png(filename=file.path(outdir,"EmptyDrops_HTO.png"), width=10, height=7, units = "in", res=300)
cowplot::plot_grid(p.barplot, p.HTO.barplot + theme(legend.position="none"), p.gradient.hb4 + theme(legend.position="none"), p.HTO.gradient.hb4 + theme(legend.position="none", axis.text.y=element_blank(), axis.title.x=element_blank()), ncol=2,  rel_heights=c(3,1), rel_widths=c(4,1), align="hv", axis="tblr", labels=c("A", "C", "B", "D"), hjust=0, vjust=1)
dev.off()

```

## COMPARE ROW SUMS FROM READS vs UMI FROM CSC.

Are reads evenly distributed among marker UMIs?

```{r}

dataCSCADTDir.reads <- "F:/Projects/ECCITE-seq/TotalSeqC_TitrationA/cite-seq-count/A1_S5_d1_ADT/read_count"

CSC.ADT.reads <- Read10X(data.dir=dataCSCADTDir.reads, gene.column=1)
CSC.ADT.reads <- CSC.ADT.reads[rownames(CSC.ADT.reads) != "unmapped",]

reads <- apply(CSC.ADT.reads,1,sum)
UMIs <- apply(CSC.ADT[,colnames(CSC.ADT.reads)],1,sum)

plotData <- data.frame(marker=gsub("(.*)\\-.*","\\1",names(reads)),reads=reads,UMIs=UMIs)

ggplot(plotData, aes(x=UMIs, y=reads)) + geom_point() + geom_smooth(method="lm") + scale_x_log10() + scale_y_log10()

reads.perCell <- as.data.frame(CSC.ADT.reads[,intersect(colnames(CSC.ADT.reads),gex.aboveInf)]) %>% mutate(marker=gsub("(.*)\\-.*","\\1",rownames(CSC.ADT.reads))) %>% pivot_longer(-marker)

UMIs.perCell <- as.data.frame(CSC.ADT[,intersect(colnames(CSC.ADT.reads),gex.aboveInf)]) %>% mutate(marker=gsub("(.*)\\-.*","\\1",rownames(CSC.ADT))) %>% pivot_longer(-marker)

plotData <- data.frame(marker=UMIs.perCell$marker,reads=reads.perCell$value,UMIs=UMIs.perCell$value)

ggplot(plotData, aes(x=UMIs, y=reads)) + geom_point(aes(color=abpanel[as.character(marker),"conc_µg_per_mL"])) + geom_smooth(method="lm") + scale_x_log10() + scale_y_log10() + scale_color_viridis_c(trans="log2", labels=scaleFUN, breaks=c(0.0375,0.15,0.625,2.5,10))

```



