---
title: "CITE-seq optimization - Cell number titration"
author: "Terkild Brink Buus"
date: "30/3/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
options(stringsAsFactors=FALSE)
```

## Load libraries etc.

```{r}
set.seed(114)
require("Seurat")
require("tidyverse")

## Load ggplot theme and defaults
source("R/ggplot_settings.R")

## Load helper functions (ggplot themes, biexp transformation etc.)
source("R/Utilities.R")

## Load predefined color schemes
source("R/color.R")

outdir <- "C:/Users/Terkild/OneDrive - Københavns Universitet/Koralovlab/ECCITE-seq/20200106 Titration 1"
```

## Load Seurat object

```{r loadSeurat}
object <- readRDS(file="data/5P-CITE-seq_Titration.rds")

object <- subset(object, subset=volume=="25µl")

Idents(object) <- object$fineCluster
table(Idents(object),as.character(object$group))
```

## Cell type and tissue overview

```{r tsnePlots}
p.tsne.number <- DimPlot(object, group.by="cellsAtStaining", reduction="tsne", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + facet_wrap(~"Volume") + scale_color_manual(values=color.number)

p.tsne.cluster <- DimPlot(object, group.by="supercluster", reduction="tsne", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + scale_color_manual(values=color.supercluster) + facet_wrap(~"Cell types")

p.tsne.finecluster <- DimPlot(object, label=TRUE, label.size=3, reduction="tsne", group.by="fineCluster", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + facet_wrap(  ~"Clusters") + guides(col=F)

p.tsne.cluster + p.tsne.finecluster + p.tsne.number
```

## Overall ADT counts

Samples stained with diluted Ab panel have reduced ADT counts.

```{r ADTcounts}
ADTcount.Abtitration <- data.frame(FetchData(object, vars=c("cellsAtStaining")), count=apply(GetAssayData(object, assay="ADT.kallisto", slot="counts"),2,sum)) %>% group_by(cellsAtStaining) %>% summarise(sum=sum(count))

p.ADTcount.Abtitration <- ggplot(ADTcount.Abtitration, aes(x=cellsAtStaining, y=sum/10^3, fill=cellsAtStaining)) + 
  geom_bar(stat="identity", col="black") + 
  scale_fill_manual(values=color.number) + 
  scale_y_continuous(expand=c(0,0,0,0.05)) + 
  guides(fill=FALSE) + 
  ylab("ADT UMI count (x10^6)") + 
  theme(panel.grid.major=element_blank(), axis.title.x=element_blank(), panel.border=element_blank(), axis.line = element_line())

p.ADTcount.Abtitration
```

## ADT sum rank by cell

Samples stained with diluted Ab panel have reduced ADT counts evenly distributed among cells

```{r ADTrank}
source("R/foot_plot.R")
p.ADTrank.Abtitration <- foot_plot(data=object$nCount_ADT.kallisto, group=object$group, linetype=object$cellsAtStaining, barcodeGroup=object$supercluster, draw.line=TRUE, draw.barcode=TRUE, draw.points=FALSE, draw.fractile=FALSE, trans="log10", barcode.stepSize = 0.1, colors=color.supercluster) + labs(linetype="Number", color="Cell type") + ylab("ADT UMI count")

p.ADTrank.Abtitration
```

## Load Ab panel annotation and concentrations

```{r loadABPanel}
abpanel <- data.frame(readxl::read_excel("data/Supplementary_Table_1.xlsx"))
rownames(abpanel) <- abpanel$Marker

markerStats <- read.table("data/markerByClusterStats.tsv")
markerStats.PBMC <- markerStats[markerStats$tissue == "PBMC",]
marker.order <- markerStats.PBMC$marker[order(-markerStats.PBMC$conc_µg_per_mL, -markerStats.PBMC$UMItotal)]
```

```{r}
ADT.matrix <- data.frame(GetAssayData(object, assay="ADT.kallisto", slot="counts"))
ADT.matrix$marker <- rownames(ADT.matrix)
ADT.matrix$conc <- abpanel[ADT.matrix$marker,"conc_µg_per_mL"]
ADT.matrix <- ADT.matrix %>% pivot_longer(c(-marker,-conc))
cell.annotation <- FetchData(object, vars=c("cellsAtStaining"))

ADT.matrix.agg <- ADT.matrix %>% group_by(cellsAtStaining=cell.annotation[name,"cellsAtStaining"], marker, conc) %>% summarise(sum=sum(value))

ADT.matrix.agg$marker.byConc <- factor(ADT.matrix.agg$marker, levels=marker.order)

ann.markerConc <- abpanel[marker.order,]
ann.markerConc$Marker <- factor(marker.order, levels=marker.order)

lines <- length(marker.order)-cumsum(sapply(split(ann.markerConc[rev(marker.order),"Marker"],ann.markerConc[rev(marker.order),"conc_µg_per_mL"]),length))+0.5
lines <- data.frame(breaks=lines[-length(lines)])

p1 <- ggplot(ann.markerConc, aes(x=1, y=Marker, fill=conc_µg_per_mL)) + 
  geom_tile(col=alpha(col="black",alpha=0.2)) + 
  geom_hline(data=lines,aes(yintercept=breaks), linetype="dashed", alpha=0.5) + 
  scale_fill_viridis_c(trans="log2") + 
  scale_x_continuous(expand=c(0,0)) + 
  labs(fill="µg/mL") + 
  theme_get() + 
  theme(axis.ticks.x=element_blank(), axis.title = element_blank(), axis.text.x=element_blank(), panel.grid=element_blank(), legend.position="right", plot.margin=unit(c(0.1,0.1,0.1,0.1),"mm"))

p2 <- ggplot(ADT.matrix.agg, aes(x=marker.byConc,y=log2(sum))) + 
  geom_line(aes(group=marker), size=1.2, color="#666666") + 
  geom_point(aes(group=cellsAtStaining, fill=cellsAtStaining), pch=21, size=0.7) + 
  geom_vline(data=lines,aes(xintercept=breaks), linetype="dashed", alpha=0.5) + 
  scale_fill_manual(values=color.number) + 
  scale_y_continuous(breaks=c(9:17)) + 
  ylab("log2(UMI sum)") + 
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), legend.position="right", legend.justification="left", legend.title.align=0, legend.key.width=unit(0.2,"cm")) + 
  coord_flip()

library(patchwork)
titration.totalCount <- p1 + guides(fill=F) + p2 + guides(fill=F) + plot_spacer() + guide_area() + plot_layout(ncol=4, widths=c(1,30,0.1), guides='collect')

titration.totalCount
```

## Overall ADT counts

Samples stained with diluted Ab panel have reduced ADT counts.

```{r ADTcountsByConc}
scaleFUN <- function(x) sprintf("%.2f", x)

p.ADTcount.AbtitrationByConc <- ggplot(ADT.matrix.agg[order(-ADT.matrix.agg$conc, -ADT.matrix.agg$sum),], aes(x=cellsAtStaining, y=sum/10^6, fill=conc)) + 
  geom_bar(stat="identity", col=alpha(col="black",alpha=0.05)) + 
  scale_fill_viridis_c(trans="log2", labels=scaleFUN, breaks=c(0.0375,0.15,0.625,2.5,10)) + 
  scale_y_continuous(expand=c(0,0,0,0.05)) + 
  labs(fill="µg/mL", y=bquote("ADT UMI counts ("~10^6~")")) + 
  guides(fill=guide_colourbar(reverse=T)) + 
  theme(panel.grid.major=element_blank(), axis.title.x=element_blank(), panel.border=element_blank(), axis.line = element_line(), legend.position="right")

p.ADTcount.AbtitrationByConc
```

Plot change by dilution. We modified our approach of using a specific fractile as this may be prone to outliers in small clusters (i.e. the 0.9 fractile of a cluster of 30 will be the #3 higest cell making it prone to outliers). We thus set a threshhold of the value to be lower than the expression of the 10th highest cell value within the cluster with highest expression.

```{r}

ADT.matrix <- data.frame(GetAssayData(object, assay="ADT.kallisto", slot="counts"))
ADT.matrix$marker <- rownames(ADT.matrix)
ADT.matrix$conc <- abpanel[ADT.matrix$marker,"conc_µg_per_mL"]
ADT.matrix <- ADT.matrix %>% pivot_longer(c(-marker,-conc))

cell.annotation <- FetchData(object, vars=c("cellsAtStaining", "fineCluster"))

ADT.matrix.agg <- ADT.matrix %>% group_by(cellsAtStaining=cell.annotation[name,"cellsAtStaining"], fineCluster=cell.annotation[name,"fineCluster"], marker, conc) %>% summarise(sum=sum(value), median=quantile(value, probs=c(0.9)), nth=nth(value))

Cluster.max <- markerStats.PBMC[,c("marker","fineCluster")]
Cluster.max$fineCluster <- factor(Cluster.max$fineCluster)

ADT.matrix.aggByClusterMax <- Cluster.max %>% left_join(ADT.matrix.agg)
ADT.matrix.aggByClusterMax$marker.byConc <- factor(ADT.matrix.aggByClusterMax$marker, levels=marker.order)

p3 <- ggplot(ADT.matrix.aggByClusterMax, aes(x=marker.byConc, y=log2(nth))) + 
  geom_line(aes(group=marker), size=1.2, color="#666666") + 
  geom_point(aes(group=cellsAtStaining, fill=cellsAtStaining), pch=21, size=0.7) + 
  geom_vline(data=lines,aes(xintercept=breaks), linetype="dashed", alpha=0.5) + 
  geom_text(aes(label=fineCluster), y=Inf, adj=1, size=1.5) + 
  scale_fill_manual(values=color.number) + 
  scale_y_continuous(breaks=c(0:11), labels=2^c(0:11), expand=c(0.05,0.5)) +
  ylab("90th percentile UMI of expressing cluster") + 
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), legend.position="right", legend.justification="left", legend.title.align=0, legend.key.width=unit(0.2,"cm")) + 
  coord_flip()
  

titration.clusterCount <- p1 + theme(legend.position="none") + p3 + theme(legend.position="none") + plot_spacer() + plot_layout(ncol=4, widths=c(1,30,0.1), guides='collect')

titration.clusterCount
```

## Titration examples

```{r}
curMarker.name <- "CD31"
curMarker.DF1conc <- abpanel[curMarker.name, "conc_µg_per_mL"]

p.CD31 <- foot_plot_density_custom(marker=paste0("adt_",curMarker.name), wrap=NULL, group="cellsAtStaining", color=color.number, legend=TRUE)
p.CD31
```

## Final plot

```{r}
library("cowplot")
A <- p.ADTcount.AbtitrationByConc + theme(legend.key.width=unit(0.3,"cm"), legend.key.height=unit(0.4,"cm"), legend.text=element_text(size=unit(5,"pt")), plot.margin=unit(c(0.3,0,0.5,0),"cm"))

D <- p.CD31 + theme(plot.margin=unit(c(0.5,0,0,0),"cm"))


B1 <- p1 + theme(text = element_text(size=10), plot.margin=unit(c(0.3,0,0,0),"cm"))
B.legend <- cowplot::get_legend(B1)
B1 <- B1 + theme(legend.position="none")
B2 <- p2
B2 <- B2 + theme(legend.position="none")
C1 <- p3 + theme(legend.position="none")


ADE <- cowplot::plot_grid(A,D,NULL, labels=c("A","D",""), label_size=panel.label_size, vjust=panel.label_vjust, hjust=panel.label_hjust, ncol=1, rel_heights = c(13,17,1.5))
BC <- cowplot::plot_grid(B1, B2, C1, nrow=1, align="h", axis="tb", labels=c("B", "", "C"), label_size=panel.label_size, vjust=panel.label_vjust, hjust=panel.label_hjust, rel_widths=c(2,10,10))

png(file=file.path(outdir,"Figure 4.png"), width=figure.width.full, height=4.5, units = figure.unit, res=figure.resolution, antialias=figure.antialias)
plot_grid(ADE, BC, nrow=1, rel_widths=c(1,4), align="v", axis="l")
dev.off()
```

## Print individual titration plots (for each marker)

For supplementary figure

```{r}

plots <- list()
markers <- intersect(abpanel[order(-abpanel$conc_µg_per_mL, abpanel$Marker),"Marker"],rownames(object[["ADT.kallisto"]]))

for(i in seq_along(markers)){
  curMarker.name <- markers[i]
  plot <- foot_plot_density_custom(marker=paste0("adt_",curMarker.name), wrap=NULL, group="cellsAtStaining", color=color.number, legend=FALSE)
  plots[[i]] <- plot
}

# a bit of a hack to get celltype legend
p.withLegend <- ggplot(data.frame(supercluster=object$supercluster),aes(color=supercluster,x=1,y=1)) + geom_point(shape=15, size=1.5)
p.legend <- cowplot::get_legend(p.withLegend + theme(legend.title=element_blank(),legend.margin=margin(0,0,0,0), legend.key.size = unit(0.15,"cm"),legend.position = c(0.98,1.1), legend.justification=c(1,1), legend.direction="horizontal"))

png(file=file.path(outdir,"Supplementary Figure 3A.png"), units=figure.unit, res=figure.resolution, width=figure.width.full, height=10, antialias=figure.antialias)
curPlots <- plot_grid(plotlist=plots[1:30],ncol=6, align="h", axis="tb")
cowplot::plot_grid(NULL, p.legend, curPlots, vjust=-0.5, hjust=panel.label_hjust, label_size=panel.label_size, ncol=1, rel_heights= c(0.5, 1.3, 70))
dev.off()
png(file=file.path(outdir,"Supplementary Figure 3B.png"), units=figure.unit, res=figure.resolution, width=figure.width.full, height=10, antialias=figure.antialias)
curPlots <- plot_grid(plotlist=plots[31:52],ncol=6, align="h", axis="tb")
cowplot::plot_grid(NULL, p.legend, curPlots, NULL, vjust=-0.5, hjust=panel.label_hjust, label_size=panel.label_size, ncol=1, rel_heights= c(0.5, 1.3, 70*(4/5),70*(1/5)))
dev.off()


```

