---
title: "CITE-seq optimization - Demux, Pre-process and downsample"
author: "Terkild Brink Buus"
date: "30/3/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set()
options(stringsAsFactors=FALSE)
```

## Load libraries etc.

```{r}
set.seed(114)
require("Seurat", quietly=T)
require("tidyverse", quietly=T)
library("Matrix", quietly=T)
library("DropletUtils", quietly=T)

driveLetter <-"F:/"
t2g.file <- file.path(driveLetter,"Projects/ECCITE-seq/TotalSeqC_TitrationA/kallisto/t2g_cellranger_CD45.txt")
kallistobusDir <- file.path(driveLetter,"Projects/ECCITE-seq/TotalSeqC_TitrationA/kallisto/gex/c1/counts_unfiltered")
kallistobusDirADT <- file.path(driveLetter,"Projects/ECCITE-seq/TotalSeqC_TitrationA/kallisto/features/A1_S5.ADT_15/counts_unfiltered")
kallistobusDirHTO <- file.path(driveLetter,"Projects/ECCITE-seq/TotalSeqC_TitrationA/kallisto/features/H1_S6.HTO_A_13/counts_unfiltered")

## Load helper functions (ggplot themes, biexp transformation etc.)
source("R/Utilities.R")

## Load predefined color schemes
source("R/color.R")
```

## Load GEX data

From kallisto-bustools output. Modified from https://github.com/Sarah145/scRNA_pre_process 

```{r loadGEX}
raw_mtx <- as(t(readMM(file.path(kallistobusDir,"cells_x_genes.mtx"))), 'CsparseMatrix') # load mtx and transpose it
rownames(raw_mtx) <- read.csv(file.path(kallistobusDir,"cells_x_genes.genes.txt"), sep = '\t', header = F)[,1] # attach genes
colnames(raw_mtx) <- read.csv(file.path(kallistobusDir,"cells_x_genes.barcodes.txt"), header = F, sep = '\t')[,1] # attach barcodes

t2g <- unique(read.csv(t2g.file, sep = '\t', header=F)[,2:3]) # load t2g file
t2g <- data.frame(t2g[,2], row.names = t2g[,1])
t2g[2533,] <- "PTPRC03"
gene_sym <- t2g[as.character(rownames(raw_mtx)),1] # get symbols for gene ids

# Which rows have same gene symbol (but different Ensembl gene id)
gene_sym.duplicated <- which(gene_sym %in% gene_sym[which(duplicated(gene_sym))])

# Which genes are have duplicated entries
gene_sym.duplicated.unique <- unique(gene_sym[gene_sym.duplicated])

# Make placeholder matrix for duplicate gene symbols
raw_mtx_dedup <- Matrix(data=0,nrow=length(gene_sym.duplicated.unique),ncol=ncol(raw_mtx))
rownames(raw_mtx_dedup) <- gene_sym.duplicated.unique
colnames(raw_mtx_dedup) <- colnames(raw_mtx)

# Combine counts from genes with same gene symbol (but different Ensembl gene id)
for(i in seq_along(gene_sym.duplicated)){
  curGene <- gene_sym[gene_sym.duplicated[i]]
  curRow <- gene_sym.duplicated.unique == curGene
  raw_mtx_dedup[curRow,] <- raw_mtx_dedup[curRow,] + raw_mtx[gene_sym.duplicated[i],]
}

# Merged combined counts duplicate gene symbol with matrix of unique gene symbol counts
raw_mtx <- raw_mtx[-gene_sym.duplicated,]
rownames(raw_mtx) <- gene_sym[-gene_sym.duplicated]
raw_mtx <- rbind(raw_mtx,raw_mtx_dedup)

tot_counts <- Matrix::colSums(raw_mtx)
summary(tot_counts)
bc_rank <- DropletUtils::barcodeRanks(raw_mtx, lower = 10)

knee_plot(bc_rank)

kallisto.GEX <- raw_mtx
```

## Load Kallisto HTO data

```{r loadHTO}
res_mat <- read_count_output(kallistobusDirHTO, name = "cells_x_genes")
dim(res_mat)
tot_counts <- Matrix::colSums(res_mat)
summary(tot_counts)
bc_rank <- DropletUtils::barcodeRanks(res_mat, lower = 10)

knee_plot(bc_rank)
kallisto.HTO <- res_mat
```

## Load Kallisto ADT data

```{r loadADT}
res_mat <- read_count_output(kallistobusDirADT, name = "cells_x_genes")
dim(res_mat)
tot_counts <- Matrix::colSums(res_mat)
summary(tot_counts)
bc_rank <- DropletUtils::barcodeRanks(res_mat, lower = 10)

knee_plot(bc_rank)
kallisto.ADT <- res_mat
```

## Demultiplex by HTO

Use Seurat MULTIseqDemux to demultiplex samples (by their hashing antibody signal = HTO)

```{r demux}
object <- CreateSeuratObject(counts = kallisto.HTO, assay="HTO.kallisto")
object <- NormalizeData(object, assay = "HTO.kallisto", normalization.method = "CLR")

## Assure the matrices are in the same barcode-space
commonDrops <- Reduce("intersect",x=list(colnames(kallisto.HTO),colnames(kallisto.ADT),colnames(kallisto.GEX)))

length(commonDrops)

object <- subset(object, cells=commonDrops)
object[["ADT.kallisto"]] <- CreateAssayObject(counts=kallisto.ADT[,commonDrops])
object[["RNA.kallisto"]] <- CreateAssayObject(counts=kallisto.GEX[,commonDrops])
Key(object[["RNA.kallisto"]]) <- "rna_"
Key(object[["ADT.kallisto"]]) <- "adt_"
Key(object[["HTO.kallisto"]]) <- "hto_"


## MULTIseqDemux seems better when using unfiltered input (including empty droplets)
object <- MULTIseqDemux(object, assay="HTO.kallisto")

RidgePlot(object, assay = "HTO.kallisto", features = rownames(object[["HTO.kallisto"]]))
table(object$MULTI_ID)

object$sampleID <- object$MULTI_ID

object <- CalculateBarcodeInflections(object,barcode.column="nCount_HTO.kallisto",group.column="sampleID",threshold.low=1000)
Seurat::BarcodeInflectionsPlot(object) + scale_x_continuous(trans="log10")

object.empty <- subset(object, subset=sampleID == "Negative")
object <- subset(object, subset=sampleID %in% c(1:6))
```

## Assign annotation to each cell

```{r annotation}
## Rename groups to meaningful names
groups <- c("PBMC_50ul_1_1000k","PBMC_50ul_4_1000k","PBMC_25ul_4_1000k","PBMC_25ul_4_200k","Lung_50ul_1_500k","Lung_50ul_4_500k","Doublet","Negative")
object$group <- object$sampleID

## Keep ordering of groups for best plotting
levels(object$group) <- groups

object$tissue <- factor(c("PBMC","PBMC","PBMC","PBMC","Lung","Lung","Doublet","Negative")[object$sampleID],levels=c("PBMC","Lung"))
object$volume <- factor(c("50µl","50µl","25µl","25µl","50µl","50µl","Doublet","Negative")[object$sampleID], levels=c("50µl","25µl","Doublet","Negative"))
object$dilution <- factor(c("DF1","DF4","DF4","DF4","DF1","DF4","Doublet","Negative")[object$sampleID], levels=c("DF1","DF4","Doublet","Negative"))
object$cellsAtStaining <- factor(c("1000k","1000k","1000k","200k","500k","500k","Doublet","Negative")[object$sampleID], levels=c("1000k","500k","200k","Doublet","Negative"))
```

## Filter dead/dying cells

Based on mitochondrial reads and number of detected genes. Cutoff set to 15% MT and at least 60 expressed genes by visual inspection.

```{r filter}
DefaultAssay(object) <- "RNA.kallisto"
object[["percent.mt"]] <- PercentageFeatureSet(object, pattern = "^MT-")
#object[["percent.mt"]][is.nan(object[["percent.mt"]][,1]),1] <- 100

cutoff.percent.mt <- 15
cutoff.nFeature <- 60
FeatureScatter(object, feature1 = "percent.mt", feature2 = "nFeature_RNA.kallisto") + ggplot2::geom_vline(xintercept=cutoff.percent.mt,linetype="dashed") + scale_y_continuous(trans="log10") + ggplot2::geom_hline(yintercept=cutoff.nFeature,linetype="dashed")

table(object[["percent.mt"]]<cutoff.percent.mt & object[["nFeature_RNA.kallisto"]]>cutoff.nFeature, object$sampleID)

object <- subset(object, subset = percent.mt <= cutoff.percent.mt & nFeature_RNA.kallisto > cutoff.nFeature)
```

## Filter doublets

Doublet rate calculated from: https://satijalab.org/costpercell at 18,000 cells and 6 multiplexed samples (3.27% after HTO demux)

```{r}
library(scDblFinder)
DefaultAssay(object) <- "RNA.kallisto"

sce <- as.SingleCellExperiment(object)
sce <- scDblFinder(sce, dbr=0.0327, samples="sampleID")

table(sce$scDblFinder.class, sce$sampleID)

identical(colnames(object),colnames(sce))

object$scDblFinder.class <- sce$scDblFinder.class
object$scDblFinder.score <- sce$scDblFinder.score
object$scDblFinder.ratio <- sce$scDblFinder.ratio
object$scDblFinder.weighted <- sce$scDblFinder.weighted
rm(sce)

object$scDblFinder.class <- factor(object$scDblFinder.class, levels=c("singlet","doublet"))
  
FeatureScatter(object, feature1 = "nCount_RNA.kallisto", feature2 = "nFeature_RNA.kallisto", group.by="scDblFinder.class", cols=c(alpha("blue",0.01),alpha("red",0.5))) + scale_y_continuous(trans="log10") + scale_x_continuous(trans="log10")

object <- subset(object, subset=scDblFinder.class=="singlet")
```


## Make DSB normalization

https://mattpm.github.io/dsb/ 

```{r dsbnorm}
#devtools::install_github(repo = 'MattPM/dsb')
library(dsb)

neg_adt_matrix = GetAssayData(object.empty, assay = "ADT.kallisto", slot = 'counts') %>% as.matrix()
positive_adt_matrix = GetAssayData(object, assay = "ADT.kallisto", slot = 'counts') %>% as.matrix()
isotypes = c("IgG2A","IgG1")

normalized_matrix_individual <- positive_adt_matrix

for(i in c(1:6)){
  curCells <- which(object$sampleID == i)
  normalized_matrix_individual[,curCells] = DSBNormalizeProtein(cell_protein_matrix = positive_adt_matrix[,curCells],
                                        empty_drop_matrix = neg_adt_matrix,
                                        use.isotype.control = TRUE,
                                        isotype.control.name.vec = isotypes)
}

object = SetAssayData(object=object, assay="ADT.kallisto", slot="data", new.data = normalized_matrix_individual)

plotData <- as.data.frame(normalized_matrix_individual)
plotData$Marker <- rownames(plotData)
plotData <- reshape2::melt(plotData, id.vars=c("Marker"))
colnames(plotData) <- c("Marker","Cell","value")
ggplot(plotData,aes(x=value,y=Marker,fill=object$group[Cell])) + ggridges::geom_density_ridges(alpha=0.5, scale=3, rel_min_height = 0.01) + xlim(-5,30) + facet_grid(~object$group[Cell])

```

## Preprocess data

Run standard Seurat preprocessing on RNA modality.

```{r preprocessRNA}

object <- NormalizeData(object)
object <- FindVariableFeatures(object)
object <- ScaleData(object)
object <- RunPCA(object, verbose = FALSE)
object <- FindNeighbors(object, dims = 1:30)
object <- FindClusters(object, resolution = 0.3)
object <- RunTSNE(object,dims=1:30)
object <- RunUMAP(object,dims=1:30)

DimPlot(object, group.by="tissue", reduction="tsne")
DimPlot(object, group.by="group", reduction="tsne")
DimPlot(object, label=TRUE, reduction="tsne")

CD45 <- grep("^PTPRC[0-9]+$",rownames(object),value=TRUE)
CD45.unexpressed <- paste0("PTPRC",c("05","06","08","11","13","14"))
CD45.expressed <- setdiff(CD45,CD45.unexpressed)
FeaturePlot(object, features=c("PTPRC"))
DoHeatmap(object, features=CD45.expressed, slot="data")
```


## Label and merge clusters into "superclusters"

To make the poulations less complex and for easier visualization, we merged the clusters into major cell types

```{r superclustering}
## LINEAGE MARKERS FOR CLUSTERLABELLING
ADTplots <- FeaturePlot(object, features=c("adt_CD1a","adt_CD3","adt_CD4","adt_CD8","adt_CD11b","adt_CD14","adt_CD19","adt_CD56","adt_HLA-DR","adt_EpCAM"), label=TRUE, reduction="tsne", min.cutoff=4, col=c("lightgrey","red"), combine=FALSE)
ADTplots <- lapply(ADTplots,FUN=function(x)x+NoLegend())
CombinePlots(ADTplots,ncol=5)
FeaturePlot(object, features=c("CD3D","TRAC"), label=TRUE, reduction="tsne", min.cutoff=1.3, col=c("lightgrey","red"), combine=FALSE)



library("dplyr")
cluster.markers <- FindAllMarkers(object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
top5 <- cluster.markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_logFC)
DoHeatmap(object, features = top5$gene, slot = "data") + NoLegend() + ggplot2::scale_fill_gradientn(colors = c("blue", "white", "red"))

## COMBINE CLUSTERS TO SUPERCLUSTERS
superclusters <- c("0"="MO/MØ/DC",
                   "1"="T/NK cells",
                   "2"="T/NK cells",
                   "3"="T/NK cells",
                   "4"="B/Plasma cells",
                   "5"="T/NK cells",
                   "6"="B/Plasma cells",
                   "7"="MO/MØ/DC",
                   "8"="MO/MØ/DC",
                   "9"="B/Plasma cells",
                   "10"="MO/MØ/DC",
                   "11"="T/NK cells",
                   "12"="Other",
                   "13"="Other",
                   "14"="B/Plasma cells",
                   "15"="Other",
                   "16"="MO/MØ/DC")

object$supercluster <- factor(superclusters[as.character(Idents(object))],levels=superclusters.levels)

DimPlot(object, group.by="supercluster", reduction="tsne")
DimPlot(object, group.by="supercluster", reduction="umap")
table <- t(table(object$supercluster,object$group))
round(table/rowSums(table)*100,2)
```

## Make fine-grained clustering
```{r fineClusters}
## SEE HOW DIFFERENT GROUPS ARE LOCATED IN PRINCIPAL COMPONENTS
#temp <- cbind(object@reductions$pca@cell.embeddings[,1:50],FetchData(object, vars=c("group","volume","dilution","tissue")))
#temp <- temp[temp$tissue=="PBMC",]

#temp2 <- temp %>% pivot_longer(c(-group,-volume,-dilution,-tissue))
#temp3 <- temp2 %>% group_by(name, group) %>% summarise(mean=median(value))
#ggplot(temp3,aes(x=name,y=mean,col=group)) + geom_point()

object <- FindClusters(object, resolution = 1.2)

object$fineCluster <- Idents(object)

DimPlot(object, reduction = "tsne", label = TRUE) + NoLegend()
DimPlot(object, reduction = "umap", label = TRUE) + NoLegend()
table(Idents(object),object$sampleID)
```

## Downsampling

To better compare UMI counts between samples, we downsample within each tissue so that each sample have the same number of cells from each fine grained cluster. Most clusters based on different tissue still have a few cells assigned to it. To avoid "expressing" clusters to be based on less than 10 cells, we remove cells belonging to clusters that have less than 10 cells within a given tissue after equal downsampling.

```{r downsample}
data.fineClusters <- FetchData(object, vars=c("tissue","sampleID","fineCluster"))

## Get number of cells in each cluster for each sample
data.fineClusters <- data.fineClusters %>% group_by(tissue, sampleID, fineCluster) %>% summarize(count=length(sampleID))

## Calculate minimum cells within each cluster for each tissue and remove clusters that are represented by less than 10 cells winin a sample of a given tissue
data.fineClusters.min <- data.fineClusters %>% group_by(tissue, fineCluster) %>% summarize(count.min=min(count)) %>% filter(count.min >= 10)

downsampled.index <- c()
for(i in 1:nrow(data.fineClusters.min)){
  curCluster <- data.fineClusters.min[i,]
  
  samples <- unique(object$sampleID[object$tissue == curCluster$tissue])
  
  for(j in seq_along(samples)){
    sample <- samples[j]
    
    ## Extract indices for cells in current sample and cluster
    cellsInCluster <- which(object$sampleID == sample & object$fineCluster == curCluster$fineCluster)
    
    ## Get random subsample according to minimum for the current cluster and tissue
    addToIndex <- cellsInCluster[sample(length(cellsInCluster),curCluster$count.min)]
    
    downsampled.index <- append(downsampled.index,addToIndex)
  }
}

table(object$sampleID,object$fineCluster)
table(object$sampleID[downsampled.index],object$fineCluster[downsampled.index])

object.downsampled <- subset(object, cells=downsampled.index[sample(length(downsampled.index),length(downsampled.index))])
table(object.downsampled$sampleID, object.downsampled$orig.ident)

DimPlot(object.downsampled, split.by="sampleID", reduction="tsne", label=TRUE , ncol=4) + NoLegend()
```

## Find "ADT expressing cluster" for each tissue

```{r ADTExpressingCluster}
abpanel <- data.frame(readxl::read_excel("data/Supplementary_Table_1.xlsx"))
rownames(abpanel) <- abpanel$Marker

ADT.matrix <- data.frame(GetAssayData(object.downsampled[,object.downsampled$volume == "50µl"], assay="ADT.kallisto", slot="counts"))
ADT.matrix <- ADT.matrix %>% mutate(marker=rownames(ADT.matrix)) %>% pivot_longer(c(-marker))
cell.annotation <- FetchData(object.downsampled, vars=c("tissue", "fineCluster", "dilution"))

ADT.matrix.agg <- ADT.matrix %>% group_by(tissue=cell.annotation[name,"tissue"], fineCluster=cell.annotation[name,"fineCluster"], marker) %>% summarise(nCells=length(value), UMIsum=sum(value), nth=nth(value), median=median(value), f90=quantile(value,probs=0.9))

marker.sum <- ADT.matrix %>% group_by(tissue=cell.annotation[name,"tissue"], dilution=cell.annotation[name,"dilution"], marker) %>% summarise(UMItotal=sum(value)) %>% filter(dilution=="DF1")
marker.sum <- marker.sum[,-2]

Cluster.max <- ADT.matrix.agg %>% group_by(marker, tissue) %>% summarize(fineCluster=fineCluster[which.max(nth)])

ADT.matrix.aggByClusterMax <- Cluster.max %>% left_join(ADT.matrix.agg) %>% left_join(marker.sum) %>% left_join(abpanel, by=c("marker"="Marker"))

write.table(ADT.matrix.aggByClusterMax,"data/markerByClusterStats.tsv")
```

## Save Seurat object

```{r save}
saveRDS(object,file="data/5P-CITE-seq_Titration_full.rds")
saveRDS(object.downsampled,file="data/5P-CITE-seq_Titration.rds")
```
