---
title: "CITE-seq optimization - Staining volume and cell number titration"
author: "Terkild Brink Buus"
date: "30/3/2020"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
options(stringsAsFactors=FALSE)
```

## Load utilities

Including libraries, plotting and color settings and custom utility functions

```{r loadLibraries, results='hide', message=FALSE, warning=FALSE}
set.seed(114)
require("Seurat", quietly=T)
require("tidyverse", quietly=T)
library("Matrix", quietly=T)
library("patchwork", quietly=T)

## Load ggplot theme and defaults
source("R/ggplot_settings.R")

## Load helper functions
source("R/Utilities.R")

## Load predefined color schemes
source("R/color.R")

## Load feature_rankplot functions
source("R/feature_rankplot.R")
source("R/feature_rankplot_hist.R")
source("R/feature_rankplot_hist_custom.R")

outdir <- "figures"
data.Seurat <- "data/5P-CITE-seq_Titration.rds"
data.abpanel <- "data/Supplementary_Table_1.xlsx"
data.markerStats <- "data/markerByClusterStats.tsv"

## Make a custom function for formatting the concentration scale
scaleFUNformat <- function(x) sprintf("%.2f", x)
```

## Load Seurat object

Subset to only focus on conditions with 200k or 1 mio cells and dilution factor 4 (thus comparing 50µl to 25µl staining volume with 1 mio or 200k PBMCs at staining).

```{r loadSeurat}
object <- readRDS(file=data.Seurat)

## Show number of cells from each sample
table(object$group)

object <- subset(object, subset=dilution == "DF4" & cellsAtStaining %in% c("200k","1000k"))
object
color.volnum <- c("50µl_1000k"="#0082c8","25µl_1000k"="#f58231","25µl_200k"="#911eb4")
shape.volnum <- c("50µl_1000k"=21,"25µl_1000k"=22,"25µl_200k"=23)
object$volnum <- factor(paste(object$volume,object$cellsAtStaining,sep="_"),levels=names(color.volnum))
```

## Load Ab panel annotation and concentrations

Marker stats is reused in other comparisons and was calculated in the end of the preprocessing vignette.

```{r loadABPanel}
abpanel <- data.frame(readxl::read_excel(data.abpanel))
rownames(abpanel) <- abpanel$Marker

## As we are only working with dilution factor 4 samples here, we want to show labels accordingly
# a bit of a hack...
abpanel$conc_µg_per_mL <- abpanel$conc_µg_per_mL/4

markerStats <- read.table(data.markerStats)
markerStats.PBMC <- markerStats[markerStats$tissue == "PBMC",]
rownames(markerStats) <- paste(markerStats$marker,markerStats$tissue,sep="_")

## Make a ordering vector ordering markers per concentration and total UMI count
marker.order <- markerStats.PBMC$marker[order(markerStats.PBMC$conc_µg_per_mL, markerStats.PBMC$UMItotal, decreasing=TRUE)]

head(abpanel)
head(markerStats)
```

## Cell type and tissue overview

Make tSNE plots colored by cell type, cluster and tissue of origin.

```{r tsnePlots, fig.height=3, fig.width=7}
p.tsne.volume <- DimPlot(object, group.by="volnum", reduction="tsne", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + facet_wrap(~"Volume") + scale_color_manual(values=color.volnum)

p.tsne.cluster <- DimPlot(object, group.by="supercluster", reduction="tsne", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + scale_color_manual(values=color.supercluster) + facet_wrap(~"Cell types")

p.tsne.finecluster <- DimPlot(object, label=TRUE, label.size=3, reduction="tsne", group.by="fineCluster", pt.size=0.1, combine=FALSE)[[1]] + theme_get() + facet_wrap(  ~"Clusters") + guides(col=F)

p.tsne.cluster + p.tsne.finecluster + p.tsne.volume
```

## Overall ADT counts

Extract UMI data and calculate UMI sum per marker within each condition.

```{r calculateUMIcountsPerMarker}
## Get the data
ADT.matrix <- data.frame(GetAssayData(object, assay="ADT.kallisto", slot="counts"))
ADT.matrix$marker <- rownames(ADT.matrix)
ADT.matrix$conc <- abpanel[ADT.matrix$marker,"conc_µg_per_mL"]
ADT.matrix <- ADT.matrix %>% pivot_longer(c(-marker,-conc))

## Get cell annotations
cell.annotation <- FetchData(object, vars=c("volnum"))

## Calculate marker sum from each dilution within both tissues
ADT.matrix.agg <- ADT.matrix %>% group_by(volume=cell.annotation[name,"volnum"], marker, conc) %>% summarise(sum=sum(value))

## Order markers by concentration
ADT.matrix.agg$marker.byConc <- factor(ADT.matrix.agg$marker, levels=marker.order)

## Extract marker annotation
ann.markerConc <- abpanel[marker.order,]
ann.markerConc$Marker <- factor(marker.order, levels=marker.order)

ADT.matrix.agg.total <- ADT.matrix.agg
```

## Plot overall ADT counts by conditions

Samples stained with diluted Ab panel have reduced ADT counts.

```{r UMIcountsPerCondition, fig.width=2.5, fig.height=2}

p.UMIcountsPerCondition <- ggplot(ADT.matrix.agg.total[order(-ADT.matrix.agg$conc, -ADT.matrix.agg$sum),], aes(x=volume, y=sum/10^6, fill=conc)) + 
  geom_bar(stat="identity", col=alpha(col="black",alpha=0.05)) + 
  scale_fill_viridis_c(trans="log2", labels=scaleFUNformat, breaks=c(0.0375,0.15,0.625,2.5,10)) + 
  scale_y_continuous(expand=c(0,0,0,0.05)) + 
  labs(fill="DF4\nµg/mL", y=bquote("ADT UMI counts ("~10^6~")")) + 
  guides(fill=guide_colourbar(reverse=T)) + 
  theme(panel.grid.major=element_blank(), axis.title.x=element_blank(), panel.border=element_blank(), axis.line = element_line(), legend.position="right")

p.UMIcountsPerCondition
```

## Compare total UMI counts per marker

Plot total UMI counts for each marker at the investigated dilution factors (DF1 vs. DF4). To ease readability, we place dashed lines between each concentration.

```{r plotUMIcountsPerMarker, fig.width=4.5, fig.height=5}
## Calculate "breaks" where concentration change.
lines <- length(marker.order)-cumsum(sapply(split(ann.markerConc$Marker,ann.markerConc$conc_µg_per_mL),length))+0.5
lines <- data.frame(breaks=lines[-length(lines)])

## Make a marker by concentration "heatmap"
p.markerByConc <- ggplot(ann.markerConc, aes(x=1, y=Marker, fill=conc_µg_per_mL)) + 
  geom_tile(col=alpha(col="black",alpha=0.2)) + 
  geom_hline(data=lines,aes(yintercept=breaks), linetype="dashed", alpha=0.5) + 
  scale_fill_viridis_c(trans="log2") + 
  labs(fill="µg/mL") + 
  theme_get() + 
  theme(axis.ticks.x=element_blank(), axis.title = element_blank(), axis.text.x=element_blank(), panel.grid=element_blank(), legend.position="right", plot.margin=unit(c(0.1,0.1,0.1,0.1),"mm")) + scale_x_continuous(expand=c(0,0))
  
## Make UMI counts per Marker plot
p.UMIcountsPerMarker <- ggplot(ADT.matrix.agg, aes(x=marker.byConc,y=log2(sum))) + 
  geom_line(aes(group=marker), size=1, color="#666666", alpha=0.5) + 
  ggbeeswarm::geom_quasirandom(aes(group=volume, fill=volume, pch=volume), size=1, dodge.width=-0.75) + 
  geom_vline(data=lines,aes(xintercept=breaks), linetype="dashed", alpha=0.5) + 
  scale_fill_manual(values=color.volnum) + 
  scale_y_continuous(breaks=c(9:17)) + 
  scale_shape_manual(values=shape.volnum) + 
  ylab("log2(UMI sum)") + 
  guides(fill=guide_legend(override.aes=list(size=1.5), ncol=1, reverse=FALSE)) + 
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), legend.position="bottom", legend.justification="left", legend.title.align=0, legend.key.width=unit(0.2,"cm"), legend.title=element_blank()) + 
  coord_flip()

## Combine plot with markerByConc annotation heatmap
plotUMIcountsPerMarker <- p.markerByConc + guides(fill=F) + p.UMIcountsPerMarker + guides(fill=F, shape=F) + plot_spacer() + guide_area() + plot_layout(ncol=4, widths=c(1,30,0.1), guides='collect')

plotUMIcountsPerMarker
```

## Compare change in UMI/cell within expressing cluster

Using a specific percentile may be prone to outliers in small clusters (i.e. the 90th percentile of a cluster of 30 will be the #3 higest cell making it prone to outliers). We thus set a threshold of the value to only be the 90th percentile if cluster contains more than 100 cells. For smaller clusters, the median is used. Expressing cluster is identified in the "preprocessing" vignette.

```{r UMIinExpressingCells, fig.width=4.5, fig.height=5}
## Get the data
ADT.matrix <- data.frame(GetAssayData(object, assay="ADT.kallisto", slot="counts"))
ADT.matrix$marker <- rownames(ADT.matrix)
ADT.matrix$conc <- abpanel[ADT.matrix$marker,"conc_µg_per_mL"]
ADT.matrix <- ADT.matrix %>% pivot_longer(c(-marker,-conc))

## Get cell annotations
cell.annotation <- FetchData(object, vars=c("volnum", "fineCluster"))

## Calculate marker statistics from each dilution within each cluster
ADT.matrix.agg <- ADT.matrix %>% group_by(volume=cell.annotation[name,"volnum"], fineCluster=cell.annotation[name,"fineCluster"], marker, conc) %>% summarise(sum=sum(value), median=quantile(value, probs=c(0.9)), nth=nth(value))
ADT.matrix.agg$tissue == "PBMC"

## Use data for the previously determined expressing cluster.
Cluster.max <- markerStats[markerStats$tissue == "PBMC",c("marker","fineCluster")]
Cluster.max$fineCluster <- factor(Cluster.max$fineCluster)

ADT.matrix.aggByClusterMax <- Cluster.max %>% left_join(ADT.matrix.agg)
ADT.matrix.aggByClusterMax$marker.byConc <- factor(ADT.matrix.aggByClusterMax$marker, levels=marker.order)

p.UMIinExpressingCells <- ggplot(ADT.matrix.aggByClusterMax, aes(x=marker.byConc, y=log2(nth))) + 
  geom_line(aes(group=marker), size=1, alpha=0.5, color="#666666") + 
  ggbeeswarm::geom_quasirandom(aes(group=volume, fill=volume, pch=volume), size=1, show.legend=FALSE, dodge.width=-0.75) + 
  geom_vline(data=lines,aes(xintercept=breaks), linetype="dashed", alpha=0.5) + 
  geom_text(aes(label=paste0(fineCluster," ")), y=Inf, adj=1, size=1.5) + 
  scale_fill_manual(values=color.volnum) + 
  scale_shape_manual(values=shape.volnum) + 
  scale_y_continuous(breaks=c(0:11), labels=2^c(0:11), expand=c(0.05,0.5)) + 
  ylab("90th percentile UMI of expressing cluster") + 
  theme(axis.title.y=element_blank(), axis.text.y=element_blank(), legend.position="right", legend.justification="left", legend.title.align=0, legend.key.width=unit(0.2,"cm")) + 
  coord_flip()

## Combine plot with markerByConc annotation heatmap
UMIinExpressingCells <- p.markerByConc + theme(legend.position="none") + p.UMIinExpressingCells + theme(legend.position="none") + plot_spacer() + plot_layout(ncol=4, widths=c(1,30,0.1), guides='collect')

UMIinExpressingCells
```

## Titration examples

Most markers are largely unaffected by reducing staining volume. However, some antibodies used at low concentrations and targeting abundant epitopes are affected, an example of such is CD31:

```{r fig.width=1.4, fig.height=2.3}
## Make helper function for plotting titration plots
titrationPlot <- function(marker, gate.PBMC=NULL, gate.Lung=NULL, y.axis=FALSE, show.gate=TRUE, legend=FALSE){
  curMarker.name <- marker
  
  ## Get antibody concentration for legends
  curMarker.DF1conc <- abpanel[curMarker.name, "conc_µg_per_mL"]
  if(show.gate==TRUE){
    ## Load gating percentages from manually set DSB thresholds
    gate <- data.frame(gate=markerStats[markerStats$marker == curMarker.name & markerStats$tissue== "PBMC",c("pct")])
    gate$gate <- 1-(gate$gate/100)
    rownames(gate) <- gate$wrap
    ## Allow manual gating
    if(!is.null(gate.PBMC)) gate <- gate.PBMC
  } else {
    gate <- NULL
  }

  p <- feature_rankplot_hist_custom(data=object, 
                                    marker=paste0("adt_",curMarker.name),      
                                    group="volnum",
                                    barcodeGroup="supercluster",
                                    conc=curMarker.DF1conc, 
                                    legend=legend, 
                                    yaxis.text=y.axis, 
                                    gates=gate,
                                    histogram.colors=color.volnum, 
                                    title=curMarker.name)
  
  return(p)
}

p.CD31 <- titrationPlot("CD31", legend=TRUE)

p.CD31
```

## Final plot

```{r figure3, fig.width=7, fig.height=4.5}
A <- p.UMIcountsPerCondition + theme(legend.key.width=unit(0.3,"cm"), 
                                     legend.key.height=unit(0.4,"cm"), 
                                     legend.text=element_text(size=unit(5,"pt")),
                                     plot.margin=unit(c(0.3,0,0,0),"cm"))

B1 <- p.markerByConc + theme(text = element_text(size=10), 
                             plot.margin=unit(c(0.3,0,1,0),"cm"),
                             legend.position="none")
B2 <- p.UMIcountsPerMarker + theme(legend.position="none")
C <- p.UMIinExpressingCells + theme(legend.position="none")

BC.legend <- cowplot::get_legend(p.UMIcountsPerMarker + 
                                   theme(legend.position="bottom", 
                                         legend.direction="horizontal", 
                                         legend.background=element_blank(), 
                                         legend.box.background=element_blank(), 
                                         legend.key=element_blank(),
                                         legend.key.height=unit(2,"mm")))

D <- p.CD31 + theme(plot.margin=unit(c(0.5,0,0,0),"cm"))

AD <- cowplot::plot_grid(A,D,NULL, 
                         ncol=1, 
                         rel_heights = c(14,16,1.5),
                         labels=c("A","D",""), 
                         label_size=panel.label_size, 
                         vjust=panel.label_vjust, 
                         hjust=panel.label_hjust)

BC <- cowplot::plot_grid(B1, B2, C, 
                         nrow=1, 
                         rel_widths=c(2,10,10), 
                         align="h", 
                         axis="tb", 
                         labels=c("B", "", "C"), 
                         label_size=panel.label_size, 
                         vjust=panel.label_vjust, 
                         hjust=panel.label_hjust)

p.final <- cowplot::ggdraw(plot_grid(AD, BC, nrow=1, rel_widths=c(1.2,4), align="v", axis="l")) + 
    cowplot::draw_plot(BC.legend,0.27,0.023,0.2,0.00001)

png(file=file.path(outdir,"Supplementary Figure S5.png"), 
    width=figure.width.full, 
    height=4.7, 
    units = figure.unit, 
    res=figure.resolution, 
    antialias=figure.antialias)

  p.final
  
dev.off()

p.final
```

## Individual titration plots

For supplementary information.

```{r suppFig1, fig.width=7, fig.height=10, eval=FALSE}
plots.columns = 6
rows.max <- 5

markers <- abpanel[rownames(object[["ADT.kallisto"]]),]
markers <- markers[order(markers$Category, markers$Marker),]

plots <- list()

## Make individual plots for each marker
for(i in 1:nrow(markers)){
  curMarker <- markers[i,]
  curMarker.name <- curMarker$Marker
  y.axis <- ifelse((i-1) %in% c(0,6,12,18,24,30,36,42,48),TRUE,FALSE)
  plots[[curMarker.name]] <- titrationPlot(curMarker.name, y.axis=y.axis)
}

# a bit of a hack to make celltype legend
p.legend <- cowplot::get_legend(ggplot(data.frame(supercluster=object$supercluster), 
                                           aes(color=supercluster,x=1,y=1)) + 
  geom_point(shape=15, size=1.5) + 
  scale_color_manual(values=color.supercluster) + 
  theme(legend.title=element_blank(), 
        legend.margin=margin(0,0,0,0), 
        legend.key.size = unit(0.15,"cm"),
        legend.position = c(0.98,1.1), 
        legend.justification=c(1,1), 
        legend.direction="horizontal"))

plots.num <- length(plots)
plots.perPage <- plots.columns*rows.max
plots.pages <- ceiling(plots.num/plots.perPage)

## Make a supplementary figure split into pages
for(i in 1:plots.pages){
  start <- (i-1)*plots.perPage+1
  end <- i*plots.perPage
  end <- min(end,plots.num)
  curPlots <- c(start:end)
  plots.rows <- ceiling(length(curPlots)/plots.columns)
  
  curPlots <- cowplot::plot_grid(plotlist=plots[curPlots],ncol=plots.columns, rel_widths=c(1.1,1,1,1,1,1), align="h", axis="tb")
  curPlots.layout <- cowplot::plot_grid(NULL, p.legend, curPlots, vjust=-0.5, hjust=panel.label_hjust, label_size=panel.label_size, ncol=1, rel_heights= c(0.5, 1.3, 70/5*plots.rows))
  
  png(file=file.path(outdir,paste0("Supplementary Figure X",LETTERS[i],".png")), 
      units=figure.unit, 
      res=figure.resolution, 
      width=figure.width.full, 
      height=(2*plots.rows),
      antialias=figure.antialias)

  print(curPlots.layout)
  
  dev.off()
  
  print(curPlots.layout)
}
```
